---
title: "chapter_3"
output: html_document
date: '2023-06-16'
---

```{r}
library(deSolve)
library(DiagrammeR)
library(ggplot2)
library(tidyverse)
```

Section 3.2
1. Fill in the unspecified units for each of the quantities in Table 3.1.

Contact Rate: exposures/fish/day
Exposure Rate: at-risk exposures/fish/day
(Fraction of at-risk exposures with an infected fish in a given day.)
Death: fish/day
Fatality Rate: deaths/sick fish
Recovering: fish/day
Recovery Time: days
Losing Resistance: fish/day
Resistance Time: days

2. The system diagram in Figure 3.1 could be modified so that a conveyor is not used for the Resistant Fish reservoir. Using this approach, the Resistance Time could still be used to dictate how long a fish would typically remain resistant. Assume that the Resistance Time is 30 days and give an equation for calculating number of fish flowing through the Losing Resistance flow under this new setup. (Hint If the Resistance Time is 30 days, what fraction of the Resistant Fish would typically lose resistance each day?)

Resistance Time = 30 days
Losing Resistance = (Resistant Fish/Resistance Time) * Resistant Fish

3. What is the difference between the approach used in Question 3 and the approach based on using conveyors?

The approach used in Question 3 expresses the Losing Resistance flow as a probability of losing resistance per day. The approach based on using conveyors expresses the Losing Resistance flow as a number of fish losing resistance per day, the value of which is dependent on the number of fish in the Resistant Fish reservoir at (t-30).

Section 3.3
4. You will find here two example purpose statements for other modeling problems. For each example, identify (1) the system to be studied, (2) the behaviors to be understood, and (3) the core questions to be addressed.
a.The purpose of this modeling effort is to understand the underlying mechanisms by which global CFC production is depleting stratospheric ozone levels, and to evaluate the long-term impact of the Montreal Protocol on those levels.

1. System: Global CFC production and stratospheric ozone levels
2. Behaviors: Depletion of stratospheric ozone levels
3. Core Questions: What are the underlying mechanisms by which global CFC production is depleting stratospheric ozone levels? What is the long-term impact of the Montreal Protocol on those levels?

b.The purpose of this modeling effort is to understand why the effluent from the local waste treatment facility in a university community exhibits a spike in biochemical oxygen demand for the 2 weeks after the end of a university holiday. We also wish to determine if this spike can be eliminated by seeding the bacteria population in the settling pond.

1. System: Effluent from the local waste treatment facility in a university community
2. Behaviors: Spike in biochemical oxygen demand for the 2 weeks after the end of a university holiday
3. Core Questions: Why does the effluent from the local waste treatment facility in a university community exhibit a spike in biochemical oxygen demand for the 2 weeks after the end of a university holiday? Can this spike be eliminated by seeding the bacteria population in the settling pond?


5. You have been hired by a solid waste management company to analyze the potential for solid waste collection in a small college community whose population is increasing at a rate of 3% per year. In particular, the company is interested in evaluating the growth of waste in this community over a 20-year period. They are also interested in determining how this growth may be affected by a recycling and reuse campaign expected to begin within the community in the fifth year. Write a purpose statement for this modeling problem.

We wish to model the growth of solid waste in a small college community over a 20-year period. We also wish to determine how this growth may be affected by a recycling and reuse campaign expected to begin within the community in the fifth year.

Section 3.4
6. Identify three possible enhancements to CHAP3a.STM to improve its structural validity. Modify the model to incorporate the enhancement that you think is most important among the three.

Infectiousness: Beta (Converter constant, between 0 and 1, Probability of infection per exposure)
Contact Rate: alpha (Converter constant, Number of at-risk exposures per fish per day, given fish density and range of travel)
Fatality Rate: mu (Converter constant, Fraction of sick fish that die)
Recovery rate: gamma (Converter constant, Days)
Resistance Time: K (Converter constant, Days)
Susceptible Fish: S (Reservoir, Number of fish susceptible to infection)
Sick Fish: I (Reservoir, Number of fish infected with the disease)
Resistant Fish: R (Reservoir, Number of fish resistant to the disease)



```{r}
sird_model <- function(time, state, parameters) {
  S <- state[1] #Susceptible Pool
  I <- state[2] #Infected Pool
  R <- state[3] #Resistant Pool
  D <- state[4] #Dead pool
  N <- S + I + R #Total Living Poulation
  
  parameters <- as.list(parameters)
  
  #comment equations
  
  
  dS <- (-parameters$beta * parameters$alpha * S * I / N) +
    parameters$k * R 
  
  dI <- parameters$beta * parameters$alpha * S * I /N -   
    parameters$gamma*I - 
    parameters$mu*I
  
  dR <- parameters$gamma*I -
    (parameters$k * R)
  
  dD <- parameters$mu*I

  return(list(c(dS, dI, dR, dD)))
}

# Set the initial conditions and parameters
initial_state <- c(S = 90, I = 10, R = 0, D = 0)
parameters <- c(alpha = 5, beta = 0.05, gamma = 0.1, mu = 0.001, k = .004)

# Set the time points for which to solve the equations
times <- seq(0, 730, by = 1)

# Solve the equations using the ode solver
primarySolution <- lsoda(y = initial_state,
                  times = times, func = sird_model, parms = parameters)

ggplot(as.data.frame(primarySolution) %>%
         pivot_longer(cols = -time)) +
  geom_line(aes(x=time, y = value, group = name)) +
  facet_wrap(~name)
```


7. Check the predictive validity of CHAP3a.STM by running it under the two sets of conditions specified in baseline behavior patterns in Figure 3.2.
a. Identify modeling conditions to use to check against the first graph Figure 3.2. Run the model for 2 years (730 days) and check the system behavior. What do you see? Are there any changes to the model that are suggested from this check?

When the infected population is 0 the model exhibits steady behavior matching the first graph in Figure 3.2.
```{r}
# Set the initial conditions and parameters
initial_state <- c(S = 100, I = 0, R = 0, D = 0) #changed I to 0
parameters <- c(beta = 0.5, gamma = 0.1, mu = 0.00, k = .004)

# Set the time points for which to solve the equations
times <- seq(0, 730, by = 1)

# Solve the equations using the ode solver
solution <- lsoda(y = initial_state,
                  times = times, func = sird_model, parms = parameters)

ggplot(as.data.frame(solution)) +
  geom_line(aes(x=time, y = S), color = "blue") +
  geom_line(aes(x=time, y = I), color = "red") +
  geom_line(aes(x=time, y = R), color = "green") +
  geom_line(aes(x=time, y = D), color = "black")
```

b. Select conditions to use to in order evaluate how well the model matches the second graph in Figure 3.2 (note that several different conditions are possible). Run the model for 730 days and check the
system behavior. What do you see?

```{r}
# Set the initial conditions and parameters
initial_state <- c(S = 990, I = 10, R = 0, D = 0)
parameters <- c(beta = 0.5, gamma = 0.2, mu = 0.0, k = .004)

# Set the time points for which to solve the equations
times <- seq(0, 730, by = 1)

# Solve the equations using the ode solver
solution <- lsoda(y = initial_state,
                  times = times, func = sird_model, parms = parameters)

#pivot and assign color to pool

ggplot(as.data.frame(solution)) +
        geom_line(aes(x=time, y = S), color = "blue") +
        geom_line(aes(x=time, y = I), color = "red") +
        geom_line(aes(x=time, y = R), color = "green") +
        geom_line(aes(x=time, y = D), color = "black")
```
c. Does the model duplicate the baseline behavior patterns?If it differs from
Figure 3.2, in what ways does it differ? Does this make physical sense, given the context of the problem? Are any modifications to the model suggested by this check?

The model does not duplicate the baseline behavior patterns, instead it has a tendency to exhibit oscillations that move towards a steady state.

8. Briefly explain the behaviors you see in Figures 3.5, 3.7, and 3.9

3.5 demonstrates the effect of a single pulse to the infected pool which changes it from 0 to 10 at day 100. 3.7 shows a step inflow, adding 10 fish per day after day 100. 3.9 shows a ramp inflow, adding 10 fish per day after day 100, increasing by 10 fish per day until day 200.

9. Open the STELLAl!l model CHAP3a.STM. Modify this model to incorporate each of the following perturbations by using the PULSE, STEp, and RAMP functions. Make sure you run only one case at a time (i.e., do not combine the cases described later). For each case, run the model for 2 years. Obsing 10 fish per day after day 100. 3.9 shows a ramp inflow, adding 10 fish per day after day 100, increasing by 10 fish per day until day 200.

a. The Contact Rate stays at a level of 2 contacts/fish/day until day 100, at which time the Contact Rate increases to 4 contacts/fish/day. (Hint Add a constant value to a STEP function).

```{r}
#run from time 1 to time 2 then time 2 to time three

initial_state <- c(S = 100, I = 10, R = 0, D = 0)
parameters <- c(alpha = 5, beta = 0.5, gamma = 0.1, mu = 0.001, k = .004)

# Set the time points for which to solve the equations
times <- seq(0, 99, by = 1)

# Solve the equations using the ode solver
solution1 <- as.data.frame(lsoda(y = initial_state,
                  times = times, func = sird_model, parms = parameters))

# Set the time points for which to solve the equations
times <- seq(99, 730, by = 1)

secondary_state <-c(S = solution1$S[100], I = solution1$I[100], R = solution1$R[100], D = solution1$D[100])
parameters <- c(alpha = 5, beta = 0.02, gamma = 0.1, mu = 0.001, k = .004)

# Solve the equations using the ode solver
solution2 <- as.data.frame(lsoda(y = secondary_state,
                  times = times, func = sird_model, parms = parameters))

solution <- bind_rows(solution1, solution2)


kicked_plot <- ggplot(solution) +
  geom_line(aes(x=time, y = S), color = "blue") +
  geom_line(aes(x=time, y = I), color = "red") +
  geom_line(aes(x=time, y = R), color = "green") +
  geom_line(aes(x=time, y = D), color = "black") +
  geom_line(data = as.data.frame(primarySolution),
            aes(x=time, y = S), color = "blue", linetype = 2) +
  geom_vline(xintercept = 99) +
  
  
kicked_plot

```

b. The Contact Rate stays level at 2 contactslfish/day until day 100, at which time the Contact Rate increases to 4 contacts/fish/day. On day 400, the Contact Rate returns to 2 contactslfish/day. (Hint Use two different STEP functions and a constant value).

```{r}
#same method as previous
initial_state <- c(S = 100, I = 10, R = 0, D = 0)
parameters <- c(beta = 0.5, gamma = 0.1, mu = 0.001, k = .004)

# Set the time points for which to solve the equations
times <- seq(0, 100, by = 1)

# Solve the equations using the ode solver
solution1 <- as.data.frame(lsoda(y = initial_state,
                                 times = times, func = sird_model, parms = parameters))

# Set the time points for which to solve the equations
times <- seq(100, 400, by = 1)

secondary_state <-c(S = solution1$S[101], I = solution1$I[101], R = solution1$R[101], D = solution1$D[101])
parameters <- c(beta = 0.02, gamma = 0.1, mu = 0.001, k = .004)

# Solve the equations using the ode solver and remove the first row
solution2 <- as.data.frame(lsoda(y = secondary_state,
                                 times = times, func = sird_model, parms = parameters)) %>%
        slice(-1)


# Set the time points for which to solve the equations
times <- seq(400, 730, by = 1)

 tertiary_state <-c(S = solution2$S[300], I = solution2$I[300], R = solution2$R[300], D = solution2$D[300])
parameters <- c(beta = 0.5, gamma = 0.1, mu = 0.001, k = .004)

# Solve the equations using the ode solver
solution3 <- as.data.frame(lsoda(y = tertiary_state,
                                 times = times, func = sird_model, parms = parameters)) %>%
        slice(-1)

solution <- full_join(solution1, solution2)
solution <- full_join(solution, solution3)

ggplot(solution) +
        geom_line(aes(x=time, y = S), color = "blue") +
        geom_line(aes(x=time, y = I), color = "red") +
        geom_line(aes(x=time, y = R), color = "green") +
        geom_line(aes(x=time, y = D), color = "black")
```

c. The Resistance Time begins at 0 and then increases by 0.5 days per day of simulation. (Hint Use the RAMP function. You do not have to specify a time value).
```{r}
sird_model2 <- function(time, state, parameters) {
  S <- state[1]
  I <- state[2]
  R <- state[3]
  D <- state[4]
  N <- S + I + R
  parameters <- as.list(parameters)
  
  resistanceTime <- parameters$k_rate * time
  
  dS <- (-parameters$beta * S * I / N) + resistanceTime * R
  dI <- parameters$beta*S*I/N - parameters$gamma*I - parameters$mu*I
  dR <- parameters$gamma*I - (resistanceTime * R)
  dD <- parameters$mu*I

  return(list(c(dS, dI, dR, dD)))
}

# Set the initial conditions and parameters
initial_state <- c(S = 100, I = 10, R = 0, D = 0)
parameters <- c(beta = 0.5, gamma = 0.1, mu = 0.001, k = .004, k_rate = 0.5)

# Set the time points for which to solve the equations
times <- seq(0, 300, by = 1)

# Solve the equations using the ode solver
solution <- lsoda(y = initial_state,
                  times = times, func = sird_model2, parms = parameters)

ggplot(as.data.frame(solution) %>%
         pivot_longer(cols = -time)) +
  geom_line(aes(x=time, y = value, group = name)) +
  facet_wrap(~name)
```



11. Use the model CHAP3b.STM and STELLA's l!l Sens; Specs to help you perform a sensitivity analysis on the remaining seven exogenous variables in the Fish Disease Model. Use a +/- 50% range for each variable. Produce a brief write-up for each variable that includes:
a. A listing of the three values you used for the sensitivity analysis.
b. A graph showing the results of the sensitivity runs for that variable.
c. A brief paragraph summarizing how the system responded to the changes in the exogenous variable and why you think the system responded that way. You should also state clearly whether you consider the variable to exert high leverage on the system, low leverage, or neither. Then state what implications your analysis has for taking corrective action against disease X.

```{r}
#sensitivity analysis
initial_state <- c(S = 100, I = 10, R = 0, D = 0)
parameters_base <- c(beta = 0.5, gamma = 0.1, mu = 0.001, k = .004)
times <- seq(0, 730, by = 1)

solve_model.fns <- function(parameters) {
        solution <- as.data.frame(lsoda(y = initial_state,
                                        times = times, func = sird_model, parms = parameters))
        return(solution)
}

#matrix of parameters with one row for each kicked parameter
parameters <- parameters_base %>%
        mutate()


```
