 ---
title: "Model 2"
output: html_notebook
date: '2023-07-10'
---


```{r}
library(deSolve)
library(DiagrammeR)
library(ggplot2)
library(tidyverse)
```

A lake ecosystem, home to many phosphorus-limited algae species, is threatened with a potential influx of inorganic phosphorus runoff from a local construction project. This phosphorus may enhance algae growth in the lake, thus reducing dissolved oxygen and increasing turbidity in the water. This will in turn create problems for the aquatic animals and bottom-dwelling plants living in the lake. A local planning authority has asked you to explore the impact on the lake's phosphorus cycle due to this potential inflow of inorganic phosphorus.

# Definitions: 

## Reservoirs:
* Live_phos = living organic phosphorus (moles P)
* Dead_phos = dead organic phosphorus (moles P)
* Inorg_phos = inorganic phosphorus (moles P)

## Rate constants:
* k_up = uptake of inorganic phosphorus (1/yr/mole)
* k_decomp = decomposition of dead organic phosphorus (1/yr)
* k_death = death of living organic phosphorus (1/yr)

## Flows:
* P_uptake(k_up * Inorg_phos * Live_phos) = uptake of inorganic phosphorus(moles P / yr)
* P_decomp(k_decomp * Dead_phos) = decomposition of dead organic phosphorus (moles P/yr)
* P_death(k_death * Live_phos) = death of living organic phosphorus (moles P / yr)



```{r}
#systems diagram
grViz("
        digraph flowchart {
        layout = neato
        overlap = false
            node [shape = rectangle]
            Inorg_phos
            Live_phos
            Dead_phos

            node [shape = doublecircle]
            P_uptake [label = 'P_uptake\n(moles/yr)']
            P_decomp [label = 'P_decomp\n(moles/yr)']
            P_death [label = 'P_death\n(moles/yr)']

            Inorg_phos -> P_uptake [arrowhead = none]
            P_uptake -> Live_phos
            Live_phos -> P_death [arrowhead = none]
            P_death -> Dead_phos
            Dead_phos -> P_decomp [arrowhead = none]
            P_decomp -> Inorg_phos
        }")


```

#Model

```{r}
#definition of rate equations
phosphorus.fns <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    dLive_phos <- k_up * Inorg_phos * Live_phos - #uptake
      k_death * Live_phos #death

    dDead_phos <- k_death * Live_phos - #death
      k_decomp * Dead_phos #decomposition

    dInorg_phos <- k_decomp * Dead_phos - #decomposition
      k_up * Inorg_phos * Live_phos #uptake

      return(list(c(dLive_phos, dDead_phos, dInorg_phos)))
  })
}

initial_state <- c(Live_phos = 0.2, Dead_phos = 1, Inorg_phos = 0.1)
parameters <- c(k_up = 2.5, k_decomp = 0.05, k_death = 0.25)

times <- seq(0, 100, by = 1) #run for 100 years

#solve and store as dataframe
solution <- ode(y = initial_state, times = times, func = phosphorus.fns, parms = parameters) %>%
  as.data.frame()

ggplot(solution, aes(x = time)) +
  geom_line(aes(y = Live_phos, colour = "Live_phos")) +
    geom_line(aes(y = Dead_phos, colour = "Dead_phos")) +
    geom_line(aes(y = Inorg_phos, colour = "Inorg_phos")) +
    labs(x = "Time (years)", y = "Phosphorus (moles P)", colour = "Legend")

```

#Sensitivity analysis

```{r}

#define base parameters and initial states
initial_state <- c(Live_phos = 0.2, Dead_phos = 1, Inorg_phos = 0.1)
base_parms <- c(k_up = 2.5, k_decomp = 0.05, k_death = 0.25)

#solve and store base solution as dataframe
base_solution <- lsoda(y = initial_state, times = times, func = phosphorus.fns, parms = base_parms) %>%
  as.data.frame()

#run model with perturb. on specified variable with designated perturb. factor
run_model <- function(kick_var.char, perturbation_factor, parms = base_parms) {
  #add if else to check if kick_var.char is in parms, if not run with kicked state
  if (kick_var.char %in% names(parms)) {
  #store base parms seperately for modificaion
  parms1 <- parms
  
  #run model with positive kick to specified parameter
  parms1[kick_var.char] <- parms1[kick_var.char] * (1 + perturbation_factor)
  solution1 <- lsoda(y = initial_state, times = times, func = phosphorus.fns,
                    parms = parms1) %>%
    as.data.frame()
  
  solution1$kick_var <- kick_var.char #store kicked parm name
  solution1$perturbation_factor <- perturbation_factor + 1 #store kick value

  parms2 <- parms #reset parms to base values
  
  #run model with positive kick to specified parameter
    parms2[kick_var.char] <- parms2[kick_var.char] * (1 - perturbation_factor)
    solution2 <- lsoda(y = initial_state, times = times, func = phosphorus.fns,
                        parms = parms2) %>%
        as.data.frame()
    
    solution2$kick_var <- kick_var.char
    solution2$perturbation_factor <- 1 - perturbation_factor
    
    #bind pushed and pulled solutions
    solution <- rbind(solution1, solution2)
  } else {
    #run model with kicked state
    initial_state1 <- initial_state
    initial_state1[kick_var.char] <- initial_state1[kick_var.char] * (1 + perturbation_factor)
    solution1 <- lsoda(y = initial_state1, times = times, func = phosphorus.fns,
                       parms = parms) %>%
      as.data.frame()

    solution1$kick_var <- kick_var.char
    solution1$perturbation_factor <- perturbation_factor + 1

    initial_state2 <- initial_state
    initial_state2[kick_var.char] <- initial_state2[kick_var.char] * (1 - perturbation_factor)
    solution2 <- lsoda(y = initial_state2, times = times, func = phosphorus.fns,
                       parms = parms) %>%
      as.data.frame()

    solution2$kick_var <- kick_var.char
    solution2$perturbation_factor <- 1 - perturbation_factor

    solution <- rbind(solution1, solution2)
  }
    
  return(solution)
}

out.df <- data.frame(kick_var2 = c("k_up", "k_decomp","Live_phos", "Dead_phos", "Inorg_phos")) %>%
  reframe(sim = run_model(kick_var.char = kick_var2, perturbation_factor = 0.5),
          .by = kick_var2) %>%
  unnest(sim) %>%
  #bind_rows(base_solution) %>% 
  mutate(perturb_factor = case_when(perturbation_factor == 0.5~"lower", perturbation_factor == 1.5~"upper", TRUE~"unknown")) %>%
  select(-perturbation_factor)%>%#^ change name of perturbation factor to prevent scoping issue
  pivot_longer(cols = c("Live_phos", "Dead_phos", "Inorg_phos"), names_to = "phos_type") %>%
  pivot_wider(names_from = perturb_factor, values_from = value)

ggplot(out.df %>%
         filter(phos_type == "Live_phos")) +
  geom_ribbon(aes(x = time, ymax = upper, ymin = lower, fill = kick_var), alpha = 0.6) +
  geom_line(aes(x = time, y = upper), color = "red") + 
  geom_line(aes(x = time, y = lower), color = "blue") +
  geom_line(data = base_solution, aes(x = time, y = Live_phos)) +
  labs(y = "Moles Live Phosphorus") +
  facet_wrap(~kick_var, scales = "free")

ggplot(out.df %>%
  filter(phos_type == "Dead_phos")) + 
  geom_ribbon(aes(x = time, ymax = upper, ymin = lower, fill = kick_var), alpha = 0.6) +
  geom_line(aes(x = time, y = upper), color = "red") + 
  geom_line(aes(x = time, y = lower), color = "blue") +
  geom_line(data = base_solution, aes(x = time, y = Dead_phos)) +
   labs(y = "Moles Dead Phosphorus") +
  facet_wrap(~kick_var, scales = "free")

ggplot(out.df %>%
  filter(phos_type == "Inorg_phos")) + 
  geom_ribbon(aes(x = time, ymax = upper, ymin = lower, fill = kick_var), alpha = 0.6) +
  geom_line(aes(x = time, y = upper), color = "red") + 
  geom_line(aes(x = time, y = lower), color = "blue") +
  geom_line(data = base_solution, aes(x = time, y = Inorg_phos)) +
   labs(y = "Moles Inorganic Phosphorus") +
  facet_wrap(~kick_var, scales = "free")

#include legends for Upper and Lower colors
```


#Conclusions

Inorganic Phosphors:
- This pool is the most stable and is only perturbed from its steady state if the uptake kinetic is modified (which could be caused by eutrophication). For this reason, the quantity of bio-matter in the lake system would need to monitered and maintained closely in order to prevent runaway behavior.

Dead Phosphorus:
- This pool is very easily disturbed from steady state. Due to the fact that oxygen shortage in the system is cause by the decomposition of dead bio-matter to inorganic matter, any modification that minimizes dead biomater would be worth investigation.
