 ---
title: "Model 1"
output: html_notebook
date: '2023-07-10'
---

```{r}
library(deSolve)
library(DiagrammeR)
library(ggplot2)
library(tidyverse)
```

Problem 4.1 You are a wildlife manager conducting a program to reintroduce red wolves into a major wildlife refuge. Although once native to the area, these wolves were eliminated decades ago by hunters, farmers, and developers. The wolves' primary source of food is deer, of which the refuge has a large population. You would like to explore how the deer and wolf populations might interact and the impacts this interaction might have on the future populations of both species.

N_deer 'N(t)' - deer population (individuals)
birth_r 'r' - deer birth rate (individuals/individual/year)
deer_cap 'K' - carrying capacity of deer (deer)
pred_eff 'c' - predator efficiency (deer/deer/wolf/year)
N_wolf - wolf population (individuals)
wolf_death 'Wd' - natural wolf death rate (wolves/individual/year)
wolf_birth_eff 'p' - wolf birth efficiency (wolf/deer/year/wolf)

```{r}
#definition of rate equations
population.fns <- function(time, state, parameters) {
with(as.list(c(state, parameters)), {
  
    dN_deer <- birth_r*(1 - N_deer/deer_cap) * N_deer -
      pred_eff * N_wolf * N_deer
    
    dN_wolf <- wolf_birth_eff * N_wolf * N_deer -
      wolf_death * N_wolf
    
    return(list(c(dN_deer, dN_wolf)))
  })
}

#define initial states and default parameters
initial_state <- c(N_deer = 500, N_wolf = 25)
parameters <- c(birth_r = 0.25, deer_cap = 4000, pred_eff = 0.005,
                wolf_death = 0.3, wolf_birth_eff = 0.001)

times <- seq(0, 100, by = 1) #run for 100 years

#solve and store as dataframe
solution <- lsoda(y = initial_state, times = times, func = population.fns,
                  parms = parameters) %>%
    as.data.frame()

#plot polulations against time to observe changes
ggplot(solution, aes(x = time)) +
  geom_line(aes(y = N_deer, color = "Deer")) +
  geom_line(aes(y = N_wolf, color = "Wolf")) +
  labs(x = "Time(years)", y = "Population", color = "Species") +
  theme_bw()

```




```{r}
#sensitivity analysis

#define base parameters and initial states
initial_state <- c(N_deer = 500, N_wolf = 25)
base_parms <- c(birth_r = 0.25, deer_cap = 4000, pred_eff = 0.005,
                wolf_death = 0.3, wolf_birth_eff = 0.001)

#solve and store base solution as dataframe
base_solution <- lsoda(y = initial_state, times = times, func = population.fns, parms = base_parms) %>%
  as.data.frame()

#run model with perturb. on specified variable with designated perturb. factor
run_model <- function(kick_var.char, perturbation_factor, parms = base_parms) {
  #add if else to check if kick_var.char is in parms, if not run with kicked state
  if (kick_var.char %in% names(parms)) {
  #store base parms seperately for modificaion
  parms1 <- parms
  
  #run model with positive kick to specified parameter
  parms1[kick_var.char] <- parms1[kick_var.char] * (1 + perturbation_factor)
  solution1 <- lsoda(y = initial_state, times = times, func = population.fns,
                    parms = parms1) %>%
    as.data.frame()
  
  solution1$kick_var <- kick_var.char #store kicked parm name
  solution1$perturbation_factor <- perturbation_factor + 1 #store kick value

  parms2 <- parms #reset parms to base values
  
  #run model with positive kick to specified parameter
    parms2[kick_var.char] <- parms2[kick_var.char] * (1 - perturbation_factor)
    solution2 <- lsoda(y = initial_state, times = times, func = population.fns,
                        parms = parms2) %>%
        as.data.frame()
    
    solution2$kick_var <- kick_var.char
    solution2$perturbation_factor <- 1 - perturbation_factor
    
    #bind pushed and pulled solutions
    solution <- rbind(solution1, solution2)
  } else {
    #run model with kicked state
    initial_state1 <- initial_state
    initial_state1[kick_var.char] <- initial_state1[kick_var.char] * (1 + perturbation_factor)
    solution1 <- lsoda(y = initial_state1, times = times, func = population.fns,
                       parms = parms) %>%
      as.data.frame()

    solution1$kick_var <- kick_var.char
    solution1$perturbation_factor <- perturbation_factor + 1

    initial_state2 <- initial_state
    initial_state2[kick_var.char] <- initial_state2[kick_var.char] * (1 - perturbation_factor)
    solution2 <- lsoda(y = initial_state2, times = times, func = population.fns,
                       parms = parms) %>%
      as.data.frame()

    solution2$kick_var <- kick_var.char
    solution2$perturbation_factor <- 1 - perturbation_factor

    solution <- rbind(solution1, solution2)
  }
    
  return(solution)
}

out.df <- data.frame(kick_var2 = c("birth_r", "deer_cap", "pred_eff", "wolf_death", "wolf_birth_eff", "N_deer", "N_wolf")) %>%
  reframe(sim = run_model(kick_var.char = kick_var2, perturbation_factor = 0.5),
          .by = kick_var2) %>%
  unnest(sim) %>%
  #bind_rows(base_solution) %>% 
  mutate(perturbation_factor = case_when(perturbation_factor == 0.5~"lower", perturbation_factor == 1.5~"upper", TRUE~"unknown"))%>%
  #^ change name of perturbation factor to prevent scoping issue
  pivot_longer(cols = c("N_deer", "N_wolf"), names_to = "species") %>%
  pivot_wider(names_from = perturbation_factor, values_from = value)

ggplot(out.df %>%
         filter(species == "N_deer")) +
  geom_ribbon(aes(x = time, ymax = upper, ymin = lower, fill = kick_var), alpha = 0.6) +
  geom_line(aes(x = time, y = upper), color = "red") + 
  geom_line(aes(x = time, y = lower), color = "blue") +
  geom_line(data = base_solution, aes(x = time, y = N_deer)) +
  labs(y = "Individuals") +
  facet_wrap(~kick_var, scales = "free")

ggplot(out.df %>%
  filter(species == "N_wolf")) + 
  geom_ribbon(aes(x = time, ymax = upper, ymin = lower, fill = kick_var), alpha = 0.6) +
  geom_line(aes(x = time, y = upper), color = "red") + 
  geom_line(aes(x = time, y = lower), color = "blue") +
  geom_line(data = base_solution, aes(x = time, y = N_wolf)) +
   labs(y = "Individuals") +
  facet_wrap(~kick_var, scales = "free")

```

Conclusion:  

Observations of the deer population
- Lowered wolf_birth_eff (the constant determining how a successful kill is translated into wolf population growth) maintained a larger and less volatile deer population. If resources can sustain this population size, this would mean deer are not over-hunted
- Increased wolf_death (natural rate of wolf death) also resulted in a larger, steadier deer population

Observations of the wolf population
- Increased predator efficiency resulted in a smaller, stabler wolf population
- Lowered wolf_birth_eff resulted in a more stable wolf population
- Increased wolf_death resulted in a more stable polulation

(discuss minimizing deer population)
